Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-30 overleaf/biometric_vectorization_example.tex
Edits to Section 5.

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_index!1 = 0
for i in range(0, N!0):
   min_sum!2 = PHI(min_sum!1, min_sum!4)
   min_index!2 = PHI(min_index!1, min_index!4)
   sum!2 = 0
   for j in range(0, D!0):
     sum!3 = PHI(sum!2, sum!4)
     d!3 = (S!0[((i * D!0) + j)] - C!0[j])
     p!3 = (d!3 * d!3)
     sum!4 = (sum!3 + p!3)
   !1!2 = (sum!3 < min_sum!2)
   min_sum!3 = sum!3
   min_index!3 = i
   min_sum!4 = MUX(!1!2, min_sum!3, min_sum!2)
   min_index!4 = MUX(!1!2, min_index!3, min_index!2)
!2!1 = (min_sum!2, min_index!2)   
\end{pythonn}
}

@@

In the code below statements (e.g., {\sf min\_sum!3 = i}) are implicitly vectorized. 
%$\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ statements have slightly different interpretation. %\ana{This is vague, may need elaboration.}
The example illustrates the two different versions of $\mathit{raise\_dim}$. {\sf raise\_dim(C!0, j, (i:N!0,j:D!0))} reshapes
the read-only input array, while {\sf drop\_dim(sum!3)} removes the $j$ dimension of {\sf sum!3}.
%\ana{More here! One just adds a least-significant dimension. The other one may reshape an input array.}

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_index!1 = 0
min_index!1^ = raise_dim(min_index!1, (i:N!0))
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))
for i in range(0, N!0):
   min_sum!2 = PHI(min_sum!1^, min_sum!4)
   min_index!2 = PHI(min_index!1^, min_index!4) 
   sum!2 = 0 // Will lift, when hoisted
   sum!2^ = raise_dim(sum!2, (j:D!0)) 
   for j in range(0, D!0):
      sum!3 = PHI(sum!2^, sum!4)
      d!3 = S!0^ - C!0^
      p!3 = (d!3 * d!3) 
      sum!4 = (sum!3 + p!3)
   sum!3^ = drop_dim(sum!3)     
   !1!2 = (sum!3^ < min_sum!2)
   min_sum!3 = sum!3^
   min_index!3 = i  // Same-level, will lift when hoisted
   min_sum!4 = MUX(!1!2, min_sum!3, min_sum!2)
   min_index!4 = MUX(!1!2, min_index!3, min_index!2)
min_sum!2^ = drop_dim(min_sum!2)
min_index!2^ = drop_dim(min_index!2)   
!2!1 = (min_sum!2^, min_index!2^)     
\end{pythonn}
}

@@

Starting at the inner phi-node {\sf sum!3 = PHI(...)}, the algorithm first computes its closure. The closure amounts to the phi-node itself and the addition node {\sf sum!4 = (sum!3 + p!3)}, accounting for the loop-carried dependency of the computation of {\sf sum}. The algorithm replaces this closure with a FOR loop on $j$ removing vectoriaztion on $j$. Note that the SUB and MUL computations remain outside of the loop as they do not depend on phi-nodes that are part of cycles. The dependences are from {\sf p!3[I,J] = (d!3[I,J] * d!3[I,J])} to the monolithic FOR loop and from the FOR loop to {\sf sum!3\^ = drop\_dim(sum!3)}. Lower case index, e.g., \texttt{i}, indicates non-vectorized dimension, while uppercase index, e.g., \texttt{I} indicates vectorized dimension.

After processing inner loop code becomes:

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_index!1 = 0
min_index!1^ = raise_dim(min_index!1, (i:N!0))
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))
for i in range(0, N!0):
  min_sum!2[I] = PHI(min_sum!1^[I], min_sum!4[I]) 
  min_index!2[I] = PHI(min_index!1^[I], min_index!4[I])  
  sum!2 = [0,..,0] 
  sum!2^ = raise_dim(sum!2, (j:D!0))
  d!3[I,J] = S!0^[I,J] - C!0^[I,J]
  p!3[I,J] = (d!3[I,J] * d!3[I,J])
  for j in range(0, D!0):
    sum!3[I,j] = PHI(sum!2^[I,j], sum!4[I,j-1])       
    sum!4[I,j] = (sum!3[I,j] + p!3[I,j])
  sum!3^ = drop_dim(sum!3)     
  !1!2[I] = (sum!3^[I] < min_sum!2[I])
  min_sum!3 = sum!3^
  min_index!3 = i
  min_sum!4[I] = MUX(!1!2[I], min_sum!3[I], min_sum!2[I])
   min_index!4[I] = MUX(!1!2[I], min_index!3[I], min_index!2[I])
min_sum!2^ = drop_dim(min_sum!2)
min_index!2^ = drop_dim(min_index!2)   
!2!1 = (min_sum!2^, min_index!2^)
\end{pythonn}
}

When processing the outer loop two closures arise, one for {\sf min\_sum!2[I] = PHI(...)} and one 
for {\sf min\_index!2[I] = PHI(...)}. Since the two closures \emph{do not} intersect, we have two distinct FOR-loops on $i$:

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_index!1 = 0
min_index!1^ = raise_dim(min_index!1, (i:N!0))
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

sum!2 = [0,..,0]
sum!2^ = raise_dim(sum!2, (j:D!0))
d!3[I,J] = S!0^[I,J] - C!0^[I,J]
p!3[I,J] = (d!3[I,J] * d!3[I,J])

for j in range(0, D!0):
  sum!3[I,j] = PHI(sum!2^[I,j], sum!4[I,j-1])       
  sum!4[I,j] = (sum!3[I,j] + p!3[I,j])

sum!3^ = drop_dim(sum!3)     
min_index!3 = [0,1,2,...N!0-1] // or min_index!3 = [i, (i:N!0)]
min_sum!3 = sum!3^

for i in range(0, N!0):
  min_sum!2[i] = PHI(min_sum!1^[i], min_sum!4[i-1]) 
  !1!2[i] = (sum!3^[i] < min_sum!2[i])
  min_sum!4[i] = MUX(!1!2[i], min_sum!3[i], min_sum!2[i])

for i in range(0, N!0):
  min_index!2[i] = PHI(min_index!1^[i], min_index!4[i-1])  
   min_index!4[i] = MUX(!1!2[i], min_index!3[i], min_index!2[i])

min_sum!2^ = drop_dim(min_sum!2)
min_index!2^ = drop_dim(min_index!2)   
!2!1 = (min_sum!2^, min_index!2^)
\end{pythonn}
}

@@

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_index!1 = 0
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

sum!2 = [0,..,0]
d!3[I,J] = S!0^[I,J] - C!0^[I,J]
p!3[I,J] = (d!3[I,J] * d!3[I,J])

// j is redundant for sum!3 and sum!4
for j in range(0, D!0):
  sum!3[I] = PHI(sum!2[I], sum!4[I])       
  sum!4[I] = (sum!3[I] + p!3[I,j])

// drop_dim is redundant, removing
// then copy propagation and dead code elimination 
min_index!3 = [0,1,2,...N!0-1] // or min_index!3 = [i, (i:N!0)]

// i is redundant for min_sum!2, min_sum!4 but not for !12![i]
for i in range(0, N!0):
  min_sum!2 = PHI(min_sum!1, min_sum!4) 
  !1!2[i] = (sum!3[i] < min_sum!2)
  min_sum!4 = MUX(!1!2[i], sum!3[i], min_sum!2)

// same, i is redundant for min_index!2, min_index!4
for i in range(0, N!0):
  min_index!2 = PHI(min_index!1, min_index!4)  
   min_index!4 = MUX(!1!2[i], min_index!3[i], min_index!2)

// drop_dim becomes redundant
!2!1 = (min_sum!2, min_index!2)
\end{pythonn}
}

@@

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_index!1 = 0
S!0^ = raise_dim(S!0, ((i * D!0) + j), (i:N!0,j:D!0))
C!0^ = raise_dim(C!0, j, (i:N!0,j:D!0))

sum!2 = [0,..,0]
d!3[I,J] = SUB_SIMD(S!0^[I,J],C!0^[I,J])
p!3[I,J] = MUL_SIMD(d!3[I,J], d!3[I,J])

for j in range(0, D!0):
  // I dim is a noop. sum is already a one-dimensional vector
  sum!3[I] = PHI(sum!2[I], sum!4[I])       
  sum!4[I] = ADD_SIMD(sum!3[I],p!3[I,j])

min_index!3 = [0,1,...N!0-1]

for i in range(0, N!0):
  min_sum!2 = PHI(min_sum!1, min_sum!4) 
  !1!2[i] = CMP(sum!3[i],min_sum!2)
  min_sum!4 = MUX(!1!2[i], sum!3[i], min_sum!2)

for i in range(0, N!0):
  min_index!2 = PHI(min_index!1, min_index!4)  
  min_index!4 = MUX(!1!2[i], min_index!3[i], min_index!2)

!2!1 = (min_sum!2, min_index!2)   
\end{pythonn}
}

>>>>

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_idx!1 = 0
for i in range(0, N):
   min_sum!2 = PHI(min_sum!1, min_sum!4)
   min_idx!2 = PHI(min_idx!1, min_idx!4)
   sum!2 = 0
   for j in range(0, D):
      sum!3 = PHI(sum!2, sum!4)
      d = SUB(S[((i * D) + j)],C[j])
      p = MUL(d,d)
      sum!4 = ADD(sum!3,p)
   t = CMP(sum!3,min_sum!2)
   min_sum!3 = sum!3
   min_idx!3 = i
   min_sum!4 = MUX(t, min_sum!3, min_sum!2)
   min_idx!4 = MUX(t, min_idx!3, min_idx!2)
return (min_sum!2, min_idx!2)

\end{pythonn}
}

@@

In the code below statements (e.g., {\sf min\_sum!3 = sum!3}) are implicitly vectorized. 
%$\mathit{raise\_dim}$ and $\mathit{drop\_dim}$ statements have slightly different interpretation. %\ana{This is vague, may need elaboration.}
The example illustrates the two different versions of $\mathit{raise\_dim}$. E.g., {\sf raise\_dim(C, j, (i:N,j:D))} reshapes
the read-only input array. {\sf drop\_dim(sum!3)} removes the $j$ dimension of {\sf sum!3}.
%\ana{More here! One just adds a least-significant dimension. The other one may reshape an input array.}

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_sum!1^ = raise_dim(min_sum!1, (i:N))
min_idx!1 = 0
min_idx!1^ = raise_dim(min_idx!1, (i:N))
S^ = raise_dim(S, ((i * D) + j), (i:N,j:D))
C^ = raise_dim(C, j, (i:N,j:D))
for i in range(0, N):
   min_sum!2 = PHI(min_sum!1^, min_sum!4)
   min_idx!2 = PHI(min_idx!1^, min_idx!4) 
   sum!2 = 0 // Will lift, when hoisted
   sum!2^ = raise_dim(sum!2, (j:D)) 
   for j in range(0, D):
      sum!3 = PHI(sum!2^, sum!4)
      d = SUB(S^,C^)
      p = MUL(d,d) 
      sum!4 = ADD(sum!3,p)
   sum!3^ = drop_dim(sum!3)     
   t = CMP(sum!3^,min_sum!2)
   min_sum!3 = sum!3^
   min_idx!3 = i  // Same-level, will lift when hoisted
   min_sum!4 = MUX(t, min_sum!3, min_sum!2)
   min_idx!4 = MUX(t, min_idx!3, min_idx!2)
min_sum!2^ = drop_dim(min_sum!2)
min_idx!2^ = drop_dim(min_idx!2)   
return (min_sum!2^, min_idx!2^)     
\end{pythonn}
}

@@

Starting at the inner phi-node {\sf sum!3 = PHI(...)}, the algorithm first computes its closure. The closure amounts to the phi-node itself and the addition node {\sf sum!4 = ADD(sum!3,p!3)}, accounting for the loop-carried dependency of the computation of {\sf sum}. The algorithm replaces this closure with a for-loop on $j$ removing vectorization on $j$. Note that the SUB and MUL computations remain outside of the loop as they do not depend on PHI-nodes that are part of cycles. The dependences are from {\sf p[I,J] = MUL(d[I,J],d[I,J])} to the monolithic for-loop and from the for-loop to {\sf sum!3\^ = drop\_dim(sum!3)}. Lower case index, e.g., \texttt{i}, indicates non-vectorized dimension, while uppercase index, e.g., \texttt{I} indicates vectorized dimension.

After processing the inner loop code becomes:

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_sum!1^ = raise_dim(min_sum!1, (i:N!0))
min_idx!1 = 0
min_idx!1^ = raise_dim(min_idx!1, (i:N))
S^ = raise_dim(S, ((i * D) + j), (i:N,j:D))
C^ = raise_dim(C, j, (i:N,j:D))
for i in range(0, N):
  min_sum!2[I] = PHI(min_sum!1^[I], min_sum!4[I]) 
  min_idx!2[I] = PHI(min_idx!1^[I], min_idx!4[I])  
  sum!2 = [0,..,0] 
  sum!2^ = raise_dim(sum!2, (j:D))
  d[I,J] = SUB(S^[I,J],C^[I,J])
  p[I,J] = MUL(d[I,J],d[I,J])
  for j in range(0, D):
    sum!3[I,j] = PHI(sum!2^[I,j], sum!4[I,j-1])       
    sum!4[I,j] = ADD(sum!3[I,j],p[I,j])
  sum!3^ = drop_dim(sum!3)     
  t[I] = CMP(sum!3^[I],min_sum!2[I])
  min_sum!3 = sum!3^
  min_idx!3 = i
  min_sum!4[I] = MUX(t[I], min_sum!3[I], min_sum!2[I])
  min_idx!4[I] = MUX(t[I], min_idx!3[I], min_idx!2[I])
min_sum!2^ = drop_dim(min_sum!2)
min_idx!2^ = drop_dim(min_idx!2)   
return (min_sum!2^, min_idx!2^)
\end{pythonn}
}

When processing the outer loop two closures arise, one for {\sf min\_sum!2[I] = PHI(...)} and one 
for {\sf min\_idx!2[I] = PHI(...)}. Since the two closures \emph{do not} intersect, we have two distinct for-loops on $i$:

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_sum!1^ = raise_dim(min_sum!1, (i:N))
min_idx!1 = 0
min_idx!1^ = raise_dim(min_idx!1, (i:N))
S^ = raise_dim(S, ((i * D) + j), (i:N,j:D))
C^ = raise_dim(C, j, (i:N,j:D))

sum!2 = [0,..,0]
sum!2^ = raise_dim(sum!2, (j:D))
d[I,J] = SUB(S^[I,J], C^[I,J])
p[I,J] = MUL(d[I,J], d[I,J])

for j in range(0, D):
  sum!3[I,j] = PHI(sum!2^[I,j], sum!4[I,j-1])       
  sum!4[I,j] = ADD(sum!3[I,j],p[I,j])

sum!3^ = drop_dim(sum!3)     
min_idx!3 = [0,1,2,...N-1] // i.e., min_idx!3 = [i, (i:N)]
min_sum!3 = sum!3^

for i in range(0, N):
  min_sum!2[i] = PHI(min_sum!1^[i], min_sum!4[i-1]) 
  t[i] = CMP(sum!3^[i], min_sum!2[i])
  min_sum!4[i] = MUX(t[i], min_sum!3[i], min_sum!2[i])

for i in range(0, N):
  min_idx!2[i] = PHI(min_idx!1^[i], min_idx!4[i-1])  
  min_idx!4[i] = MUX(t[i], min_idx!3[i], min_idx!2[i])

min_sum!2^ = drop_dim(min_sum!2)
min_idx!2^ = drop_dim(min_idx!2)   
return (min_sum!2^, min_idx!2^)
\end{pythonn}
}

@@

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_idx!1 = 0
S^ = raise_dim(S, ((i * D) + j), (i:N,j:D))
C^ = raise_dim(C, j, (i:N,j:D))

sum!2 = [0,..,0]
d[I,J] = SUB(S^[I,J],C^[I,J])
p[I,J] = MUL(d[I,J],d[I,J])

// j is redundant for sum!3 and sum!4
for j in range(0, D):
  sum!3[I] = PHI(sum!2[I], sum!4[I])       
  sum!4[I] = ADD(sum!3[I], p[I,j])

// drop_dim is redundant, removing
// then copy propagation and dead code elimination 
min_idx!3 = [0,1,2,...N-1] // i.e., min_idx!3 = [i, (i:N)]

// i is redundant for min_sum!2, min_sum!4 but not for t[i]
for i in range(0, N):
  min_sum!2 = PHI(min_sum!1, min_sum!4) 
  t[i] = CMP(sum!3[i],min_sum!2)
  min_sum!4 = MUX(t[i], sum!3[i], min_sum!2)

// same, i is redundant for min_idx!2, min_idx!4
for i in range(0, N):
  min_idx!2 = PHI(min_idx!1, min_idx!4)  
  min_idx!4 = MUX(t[i], min_idx!3[i], min_idx!2)

// drop_dim becomes redundant
return (min_sum!2, min_idx!2)
\end{pythonn}
}

@@

{\small
\begin{pythonn}
min_sum!1 = MAX_INT
min_idx!1 = 0
S^ = raise_dim(S, ((i * D) + j), (i:N,j:D))
C^ = raise_dim(C, j, (i:N,j:D))

sum!2 = [0,..,0]
d[I,J] = SUB_SIMD(S^[I,J],C^[I,J])
p[I,J] = MUL_SIMD(d[I,J], d[I,J])

for j in range(0, D):
  // I dim is a noop. sum is already a one-dimensional vector
  sum!3[I] = PHI(sum!2[I], sum!4[I])       
  sum!4[I] = ADD_SIMD(sum!3[I],p[I,j])

min_idx!3 = [0,1,...N-1]

for i in range(0, N):
  min_sum!2 = PHI(min_sum!1, min_sum!4) 
  t[i] = CMP(sum!3[i],min_sum!2)
  min_sum!4 = MUX(t[i], sum!3[i], min_sum!2)

for i in range(0, N):
  min_idx!2 = PHI(min_idx!1, min_idx!4)  
  min_idx!4 = MUX(t[i], min_idx!3[i], min_idx!2)

return (min_sum!2, min_idx!2)   
\end{pythonn}
}
